Open-source tools in R for landscape ecology

### 2021-01-27 paper/paper.Rmd
Small edits in the abstract.

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- R packages
- reproducibility
- scientific software
abstract: |
  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. **Recent findings** In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. **Summary** In this we find that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually updating the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

API & Application programming interface. A set of protocols for sending and retrieving information from a server \\
cell & Smallest, rectangular unit of raster data model \\
CRAN & Comprehensive R Archive Network. A repository for *R* package \\
CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
EPSG & Public registry of geodetic datums, spatial reference systems, Earth ellipsoids, coordinate transformations and related units of measurement \\
GDAL & Open source translator library for raster and vector geospatial data formats \\
GIS & Geographical information system \\
landscape & Mosaic of land covers, ecosystems, habitat types, land uses \\
NLM & Neutral landscape model \\
open-source software & Software released under licenses that allow to use, modify and distribute it \\
patch & Neighboring cells with same value \\
PROJ & Open source generic coordinate transformation software \\
raster & Cell grid based data model \\
resolution & Size in meter or degree of one cell \\
*R* package & User-created software extension to *R* programming language \\
SDM & Species distribution modelling \\
Simple Features & A set of standards that specify a common storage and access model of geographic features \\
vector & Points, lines, and polygons-based data model \\

>>>>

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- R packages
- reproducibility
- scientific software
abstract: |
  **Purpose of review** Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the *R* programming language provides a wealth of community developed tools for landscape ecology. **Recent findings** In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of *R* users within the landscape ecology community. **Summary** We found that landscape ecologists are generally satisfied with the functionality available within *R*, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

API & Application programming interface. A set of protocols for sending and retrieving information from a server \\
cell & Smallest, rectangular unit of raster data model \\
CRAN & Comprehensive R Archive Network. A repository for *R* package \\
CRS & Coordinate references systems. A coordinate-based local, regional or global system used to locate geographical entities \\
EPSG & Public registry of geodetic datums, spatial reference systems, Earth ellipsoids, coordinate transformations and related units of measurement \\
GDAL & Open source translator library for raster and vector geospatial data formats \\
GIS & Geographical information system \\
landscape & Mosaic of land covers, ecosystems, habitat types, land uses \\
NLM & Neutral landscape model \\
open-source software & Software released under licenses that allow to use, modify and distribute it \\
patch & Neighboring cells with same value \\
PROJ & Open source generic coordinate transformation software \\
raster & Cell grid based data model \\
resolution & Size in meter or degree of one cell \\
*R* package & User-created software extension to *R* programming language \\
SDM & Species distribution modeling \\
Simple Features & A set of standards that specify a common storage and access model of geographic features \\
vector & Points, lines, and polygons-based data model \\

### 2021-01-27 paper/paper.tex
Small edits in the abstract.

\begin{abstract}
\textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the \emph{R} programming language provides a wealth of community developed tools for landscape ecology. \textbf{Recent findings} In this paper, we outline existing packages for the downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of \emph{R} users within the landscape ecology community. \textbf{Summary} In this we find that landscape ecologists are generally satisfied with the functionality available within \emph{R}, and that as a community they are continually updating the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
\\
\keywords{
        spatial data \and
        statistical programming language \and
        R packages \and
        reproducibility \and
        scientific software \and
    }

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The \emph{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \emph{R} packages directly depend or imports \emph{sp}.
However, \emph{sp} is not actively developed anymore, and its recommended successor is the \emph{sf} package \cite{Pebesma2018}.
Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019a}.
\emph{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

To better understand how the landscape ecology community uses \emph{R}, we conducted a short survey using mailing lists and social media to reach the community.
In total, the survey was answered by 109 participants, of which the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (14.7\%).
Other, less frequent answers were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).

@@

We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig. \ref{fig:fig-survey} A)).

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').

@@

Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig. \ref{fig:fig-survey} D)).

>>>>

\begin{abstract}
\textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the \emph{R} programming language provides a wealth of community developed tools for landscape ecology. \textbf{Recent findings} In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of \emph{R} users within the landscape ecology community. \textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \emph{R}, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools.
\\
\keywords{
        spatial data \and
        statistical programming language \and
        R packages \and
        reproducibility \and
        scientific software \and
    }

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The \emph{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \emph{R} packages directly depend or imports \emph{sp}.
However, \emph{sp} is not actively developed anymore, and its recommended successor is the \emph{sf} package \cite{Pebesma2018}.
Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019a}.
\emph{sf} builds on the idea of ``simple features'', a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

To better understand how the landscape ecology community uses \emph{R}, we conducted a short survey using mailing lists and social media to reach the community.
In total, the survey was answered by 109 participants, of which the majority were either ``PhD students'' (33\%), followed by ``Post-Docs'' (26.6\%) and ``Professors'' (14.7\%).
Other, less frequent answers were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).

@@

We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'', and ``nature conservation'' (Fig. \ref{fig:fig-survey} A)).

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').

@@

Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 90.83\%) and only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 9.17\%; Fig. \ref{fig:fig-survey} D)).
